\begin{thebibliography}{99}
\markboth{Bibliography}{Bibliography}


\bibitem{Aktug2007}
Aktug, I. and Naliuka, K. (2007). ConSpec: A Formal Language for Policy Specification, \emph{Proceedings of First International Workshop on Formal Languages and Analysis of Contract-Oriented Software}, pp. 107-–109.

\bibitem{Alur1990}
Alur, R. and Dill, D.L. (1990). Automata For Modeling Real-Time Systems, \emph{Proceedings of 17th International Colloquium on Automata, Languages, and Programming}, pp. 322--335.

\bibitem{Alur1994}
Alur, R. and Dill, D.L. (1994). A Theory of Timed Automata, \emph{Theoretical Computer Science} \textbf{126}, pp. 183--235.

\bibitem{Ambler2005}
Ambler, S.W. (2005). The Elements of UML(TM) 2.0 Style (Cambridge University Press, New York).

\bibitem{Andersen2006}
Andersen, J., Elsborg, E., Henglein, F., Simonsen, J.G. and Stefansen, C. (2006). Compositional specification of commercial contracts, \emph{International Journal on Software Tools for Technology Transfer} \textbf{8}, pp. 485-–516.

\bibitem{Andres2009}
Andr\'{e}s, C., D\'{\i}az, G., Mart\'{\i}nez, E. and Zhang, Y. (2009). Formal Study of Prioritized Service Compositions, \emph{Proceedings of Fifth International Conference on Signal Image Technology and Internet Based Systems}, pp. 355-–362.

\bibitem{Barnett2005}
Barnett, M., Leino, K.R.M. and Schulte, W. (2005). The Spec\# programming system: An overview., \emph{Proceedings of Construction and Analysis of Safe, Secure, and Interoperable Smart devices} \textbf{3362}, pp. 49--69.

\bibitem{Barros2005}
Barros, A., Dumas, M. and Oaks, P. (2005). A Critical Overview of the Web Services Choreography Description Language (WS-CDL), \emph{BPTrends Newsletter} \textbf{3}.

\bibitem{Bartlett1969}
Bartlett, K.A., Scantlebury, R.A. and Wilkinson, P.T. (1969). A note on reliable full-duplex transmission over half-duplex links, \emph{Communications of the ACM} \textbf{12}, pp. 260--261.

\bibitem{Beek2003}
ter Beek, M.H., Clarence, A., Ellis, J.K. and Rozenberg, G. (2003). Synchronizations in Team Automata for Groupware Systems, \emph{Computer Supported Cooperative Work} \textbf{12}, pp. 21--69.

\bibitem{Belhaouari2007}
Belhaouari, H. and Peschanski, F. (2007). An Integrated Platform for Contract-oriented Development, \emph{Proceedings of First International Workshop on Formal Languages and Analysis of Contract-Oriented Software}, pp. 110--112.

\bibitem{Bengtsson2003}
Bengtsson, J. and Yi, W. (2003). Timed automata: Semantics, algorithms and tools, \emph{Lecture Notes on Concurrency and PetriNets} \textbf{3098}, pp. 87--124.

\bibitem{Behrmann2004}
Behrmann, G., David, A. and Larsen, K.~G. (2004). A tutorial on Uppaal, \emph{Formal Methods for the Design of Real-Time Systems} \textbf{3185}, pp. 200--236.	

\bibitem{Berners-Lee2001}
Berners-Lee, T., Hendler, J. and Lassila, O. (2001). The Semantic Web, \emph{Scientific American} \textbf{284}, pp. 34--43.

\bibitem{Beugnard1999}
Beugnard, A., Jezequel, J.M., Plouzeau, N. and Watkins, D. (1999). Making Components Contract Aware, \emph{IEEE Computer} \textbf{32}, pp. 38--45.

\bibitem{Boyer1983}
Boyer, R.S. and Moore, J.S. (1983). Proof-checking, theorem-proving and program verification, \emph{Technical Report 35}, Institute for Computing Science and Computer Applications, University of Texas at Austin

\bibitem{Bravetti2007}
Bravetti, M. and Zavattaro, G. (2007). Towards a unifying theory for choreography conformance and contract compliance, \emph{Proceedings of the Sixth International Symposium on Software Composition}, pp. 34-–50.

\bibitem{Brendan1998}
Brendan, M. and Dong, J.S. (1998). Blending Object-Z and Timed CSP: An Introduction to TCOZ, \emph{Proceedings of the 20th international conference on Software engineering}, pp. 95-–104.

%\bibitem{Brooke1996}
%Brooke, J. (1996). SUS - A ``quick and dirty'' usability scale, \emph{Usability Evaluation in Industry}, pp. 189--194.

\bibitem{Buscemi2008}
Buscemi, M.G. and Montanari. U. (2008). Cc-Pi: A Constraint-Based Language for Contracts with Service Level Agreements, \emph{Proceedings of Second International Workshop on Formal Languages and Analysis of Contract-Oriented Software}, pp. 1-–8.

\bibitem{Cambronero2006}
Cambronero, M.E., Valero, V., Aparicio, I., D\'{\i}az, G. and Pardo, J.J. (2006). Translation of UML Diagrams into Web Services Business Process Execution Language (WS-BPEL), \emph{Jornadas Cient\'{\i}fico-T\'{e}cnicas en Servicios Web (Spanish conference on Web Services)}, pp. 99--106.

\bibitem{Cambronero2007}
Cambronero, M.E. (2007). Description and Verification of Multimedia Systems and Web Services with Time Constraints, \emph{PhD. Thesis, University of Castilla-La Mancha}

\bibitem{Cambronero2007-2}
Cambronero, M.E., D\'{\i}az, G., Pardo, J.J., Valero, V. and Cuartero, F.(2007). Using UML Diagrams to Model Real-Time Web Services, \emph{Proceedings of the Second International Conference on Internet and Web Applications and Services}, p. 24.

\bibitem{Cambronero2009}
Cambronero, M.E., Valero, V. and D\'{\i}az, G. (2009). Verification of real-time systems design, \emph{Software Testing, Verification and Reliability} \textbf{20}, pp. 3--37.

\bibitem{Carbone2006}
Carbone, M., Honda, K. and Yoshida, N. (2006). A Theoretical Basis of Communication-Centres Concurrent Programming, \emph{First International Summer School on Emerging Trends in Concurrency}

\bibitem{Carpineti2006}
Carpineti, S., Castagna, G., Laneve, C. and Padovani, L. (2006). A formal account of contracts for Web services, \emph{Proceedings of Third International Workshop on Web Services and Formal Methods}, pp. 148-–162.

\bibitem{Clarke1999}
Clarke, E.M., Grumberg, J.O. and Peled, D.A. (1999). Model Checking (MIT Press).

\bibitem{Cleaveland1993}
Cleaveland, R., Parrow, J. and Steffen, B. (1993). The Concurrency Workbench: A Semantics-Based Tool for the Verification of Concurrent Systems, \emph{ACM Transactions on Programming Languages and Systems} \textbf{15}, pp. 36--72.

\bibitem{Davulcu2004}
Davulcu, H., Kifer, M. and Ramakrishnan, I.V. (2004). $\cal CTR-S$: A Logic for Specifying Contracts in Semantic Web Services, \emph{Proceedings of the Thirteenth international World Wide Web conference}, pp. 144-–153.

\bibitem{Dellarocas2000}
Dellarocas, D. (2000). Contractual Agent Societies: Negotiated Shared Context and Social Control in open Multi-agent Systems, \emph{Proceedings of Workshop on Norms and Institutions in Multi-Agent Systems, Autonomous Agents-2000}.

\bibitem{Decker2009}
Cambronero, G. (2009). Design and Analysis of Process Choreographies, \emph{PhD. Thesis, Hasso Plattner Institute, Uiversity of Postdam}

\bibitem{Desai2008}
Desai, N., Narendra, N.C. and Singh, M.P. (2008). Checking correctness of business contracts via commitments, \emph{Proceedings of 7th on Autonomous Agents and Multiagent Systems}, pp. 787–-794. 

\bibitem{Diaz2009}
D\'{\i}az, G. and Rodr\'{\i}guez, I. (2009). Automatically Deriving Choreography-Conforming Systems of Services, \emph{Proceedings of 2009 IEEE International Conference on Services Computing}, pp. 9-–16.

\bibitem{Dong2006}
Dong, J.S., Liu, Y., Sun, J. and Zhang, X. (2006). Verification of Computation Orchestration Via Timed Automata, \emph{Proceedings of Eighth International Conference on Formal Engineering Methods} \textbf{4260}, pp. 226--245.

\bibitem{Enseling2001}
Enseling, O. (2001). iContract: Design by Contract in Java, \url{http://www.javaworld.com/javaworld/jw-02-2001/jw-0216-cooltools.html}.

\bibitem{Filliâtre2008}
Filli\^atre, J.C., Baudin, P. and March\'e, C. (2008). ACSL: ANSI/ISO C Specification Language, \url{http://frama-c.cea.fr/acsl.html}.

\bibitem{Foster2007}
Foster, H., Emmerich, W., Kramer, J., Magee, J., Rosenblum, D.S. and Uchitel, S. (2007). Model Checking Service Compositions Under Resource Constraints, \emph{Proceedings of Sixth Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT International Symposium on Foundations of Software Engineering}, pp. 225--234.

\bibitem{Glabbeek1987}
van Glabbeek, R.J. (1987). Bounded nondeterminism and the approximation induction principle in process algebra, \emph{Proceedings of 4th Annual Symposium on Theoretical Aspects of Computer Science}, pp. 336--347.

\bibitem{Governatori2006}
Governatori, G., Milosevic, Z. and Sadiq, S. (2006). Compliance checking between business processes and business contracts, \emph{Proceedings of the 10th IEEE Conference on Enterprise Distributed Object Computing}, pp. 221--232.

\bibitem{Grüner2008}
Grüner, H.P. (2008). Protocol Design and (De-)Centralization, \emph{University of Mannheim and CEPR, London}.

\bibitem{Guttag1982}
Guttag, J.V., Horning, J.J. and Wing, J.M. (1982). Some Remarks on Putting Formal Specifications to Productive Use, \emph{Science of Computer Programming} \textbf{2}, pp. 53--68.

\bibitem{Hamadi2003}
Hamadi, R. and Benatallah, B. (2003). A Petri Net-based Model for Web Service Composition, \emph{Proceedings of the 14th Australasian database conference on Database technologies}, pp. 191--200.

\bibitem{Hatcliff2009}
Hatcliff, J., Leavens, G.T., Leino, k.R.M., Müller, P. and Parkinson, M. (2009). Behavioral Interface Specification Languages, \emph{Technical Report CS-TR-09-01}, School of EECS, University of Central Florida.

\bibitem{Holzmann1997}
Holzmann, G. J. (1997). The Model Checker Spin, \emph{IEEE Transactions on Software Engineering} \textbf{23}, pp. 279--295.

\bibitem{IBM2003}
IBM (2003). Web Service Level Agreement (WSLA) Language Specification, \url{http://www.research.ibm.com/wsla/WSLASpecV1-20030128.pdf}.

\bibitem{Jensen1990}
Jensen, k. (1990). Coloured Petri Nets: A high-level Language for System Design and Analysis, \emph{Lecture Notes in Computer Science} \textbf{483} (Springer, Verlag).

\bibitem{Jensen1991}
Jensen, k., Christensen, S., Huber, P. and Holla, M. (1991). Design/CPN. A Reference Manual, \emph{Meta Software Cooperation} 

\bibitem{Jensen2007}
Jensen, k., Kristensen, L.M. and Wells, L. (2007). Coloured Petri Nets and CPN Tools for modelling and validation of concurrent systems, \emph{International Journal on Software Tools for Technology Transfer} \textbf{9}, pp. 213--254.

\bibitem{Karp2000}
Karp, A.H. (2000). E-speak Explained, \emph{Technical Report HPL-2000-101}, Hewllet-Packard Laboratories.

\bibitem{Katoen1998}
Katoen, J.P. (1998). Concepts, Algorithms, and Tools for Model Checking, \emph{Lecture Notes of the course Mechanized Validation of Parallel Systems (course number 10359)}.

\bibitem{Kaynar2006}
Kaynar, D. (2006). E-speak Explained, \emph{Theory of Timed I/O Automata}, Morgan Claypool.

\bibitem{Keller2002}
Keller, A. and Ludwig, H. (2002). The WSLA Framework: Specifying and Monitoring of Service Level Agreements for Web Services, \emph{IBM research report RC22456}.

\bibitem{Krishna2008}
Krishna, P.R. and Karlapalem, K. (2008). Electronic Contracts, \emph{IEEE Internet Computing} \textbf{12}, pp. 60--68.

\bibitem{Lamsweerde1993}
van Lamsweerde, A., Dardenne, A. and Fickas, S. (1993). Goal-directed requirements acquisition, \emph{Selected Papers of the Sixth International Workshop on Software Specification and Design. Science of Computer Programming} \textbf{20}, pp. 3--50.

\bibitem{Larsen1997}
Larsen, K.G., Pettersson, P. and Yi W. (1997). UPPAAL in a Nutshell, \emph{STTT: International Journal on Software Tools for Technlogy Transfer} \textbf{1--2}, pp. 134--152.

\bibitem{Leavens1999}
Leavens, G.T., Baker, A.L. and Ruby, C. (1999). Preliminary design of JML: A behavioral interface specification language for Java, \emph{Technical Report 98-06f} \textbf{12}, Department of Computer Science, Iowa State University.

\bibitem{Linz2006}
Linz, P. (2006). An introduction to formal languages and automata, \emph{Jones \& Bartlett Learning}.

\bibitem{Lomuscio2006}
Lomuscio, A. and Raimondi, F. (2006). MCMAS: A model checker for multi-agent systems, \emph{Proceedings of 12th International Conference on Tools and Algorithms for the Construction and Analysis of System} \textbf{3920}, pp. 450--454.

\bibitem{Lomuscio2008}
Lomuscio, A., Qu, H. and Solanki, M. (2008). Towards verifying contract regulated service composition, \emph{Proceedings of IEEE International Conference on Web Services}, 254--261.

\bibitem{McIlraith2001}
McIlraith, S.A., Son, T.C. and Zeng, H. (2001). Semantic Web Services, \emph{IEEE Intelligent Systems} \textbf{16}, pp. 46-–53.

\bibitem{McNamara2006}
McNamara, P. (2006). Deontic Logic, \emph{Handbook of the History of Logic} \textbf{7}, pp. 197--289.

\bibitem{Medvidovic2000}
Medvidovic, N. and Taylon, R.N. (2000). A classification and comparison framework for software architecture description languages, \emph{IEEE Transactions on Software Engineering} \textbf{26}, pp. 70--93.

\bibitem{Meyer1992}
Meyer, B. (1992). Applying Design by Contract, \emph{IEEE Computer} \textbf{25}, pp. 40--51.

\bibitem{Meyer1992-2}
Meyer, B. (1992). Eiffel: The Language, \emph{Prentice Hall}.

\bibitem{Mills2004}
Mills, C. (2004). Using Design by Contract in C, \url{http://onlamp.com/pub/a/onlamp/2004/10/28/design_by_contract_in_c.html}.

\bibitem{Milner1980}
Milner, R. (1980). A Calculus of Communicating Systems, \emph{Lecture Notes in Computer Science} \textbf{92} (Springer, New York).

\bibitem{Milner1992}
Milner, R., Parrow, J. and Walker, D. (1992). A calculus of mobile processes I \& II, \emph{Information and Computation} \textbf{100}, pp. 1--77.

\bibitem{Moody2009}
Moody, D.L. (2009). The ``Physics'' of Notations: Toward a Scientific Basis for Constructing Visual Notations in Software Engineering, \emph{IEEE Transactions on Software Engineering} \textbf{35}, pp. 756--779.

\bibitem{Nakajima2005}
Nakajima, S. (2005). Model-checking Behavioral Specification of BPEL Applications, \emph{Proceedings of the International Workshop on Web Languages and Formal Methods} \textbf{151}, pp. 89--105.

Proceedings of the International Workshop on Web Languages and Formal Methods

\bibitem{OASIS2004}
OASIS (2004). UDDI Version 3.0.2, \url{http://uddi.org/pubs/uddi-v3.0.2-20041019.htm}.

\bibitem{OASIS2007}
OASIS (2007). Web Services Business Process Execution Language Version 2.0, \url{http://docs.oasis-open.org/wsbpel/2.0/OS/wsbpel-v2.0-OS.html}.

\bibitem{Okika2007}
Okika, J.C. and Ravn A.P. (2007). Compositionality and Compatibility of Service Contracts, \emph{Proceedings of First International Workshop on Formal Languages and Analysis of Contract-Oriented Software} \textbf{366}, pp. 101--103.

\bibitem{OMG1998}
OMG (1998). UML 2.0 XML Metadata Interchange (XMI), \emph{OMG document ad/98-10-05}.

\bibitem{OMG2003}
OMG (2003). UML 2.0 Superstructure proposal v.2.0., \emph{OMG document ad/03-01-02}.

\bibitem{Pace2009}
Pace, G.J. and Schneider, G. (2009). Challenges in the specification of full contracts, \emph{Proceedings of 7th International Conference on integrated Formal Methods}, pp. 292--306.

\bibitem{Papazoglou2007}
Papazoglou, M.P., Traverso, P., Dustdar, S. and Leymann, F. (2007). Service-Oriented Computing: State of the Art and Research Challenges, \emph{IEEE Computer} \textbf{40}, pp. 38--45.

\bibitem{Peled2001}
Peled, D. (2001). Software Reliability Methods (Springer, Berlin).

\bibitem{Peltz2003}
Peltz, C. (2003). Web Services Orchestration and Choreography, \emph{IEEE Computer} \textbf{36}, pp. 46--52.

\bibitem{Petri2008}
Petri, C.A. and Reisig W. (2008). Petri Net, \emph{Scholarpedia}.

\bibitem{Prakken1996}
Prakken, H. and Sergot M. (1996). Contrary-to-duty obligations, \emph{Studia Logica} \textbf{57}, pp. 91--115.

\bibitem{Prisacariu2008}
Prisacariu, C. and Schneider, G. (2008). A Formal Language for Electronic Contracts, \emph{Proceedings of Nineth IFIP International Conference on Formal Methods for Open Object-Based Distributed Systems}, pp. 174-–189.

\bibitem{Pu2006}
Pu, G., Zhao, X., Wang, S. and Qiu, Z. (2006). Towards the Semantics and Verification of BPEL4WS, \emph{Proceedings of the International Workshop on Web Languages and Formal Methods} \textbf{151}, pp. 33-–52.

\bibitem{Qian2007}
Qian, Y., Yuming, X., Wang, Z., Pu, G., Zhu, H. and Cai, C. (2007). Tool Support for BPEL Verification in ActiveBPEL Engine, \emph{Proceedings of 18th Australian Software Engineering Conference}, pp. 90-–100.

\bibitem{Solaiman2003}
Solaiman, E., Molina-Jimenez, C. and Shrivastava, S. (2003). Model Checking Correctness Properties of Electronic Contracts, \emph{Proceedings of First International Conference on Service Oriented Computing}, pp. 303-–318.

\bibitem{Sun1999}
Sun Microsystems (1999). Jini Architecture Specification, \url{http://www.cs.princeton.edu/courses/archive/fall99/cs597b/docs/jcpdoc1_0/specs/jini-spec/jini-spec.pdf}.

\bibitem{Vaculin2008}
Vacul\'{\i}n, R.,  Wiesner, K. and Sycara, K. (2008). Exception handling and recovery of semantic web services, \emph{Proceedings of Fourth International Conference on Networking and Services}, pp. 217-–222.

\bibitem{Valero2009}
Valero, V., Maci\`a, H. and Mart\'{\i}nez, E. (2009). Transforming WS-CDL specifications into Coloured Petri Nets, \emph{Proceedings of International Workshop on Petri Nets and Software Engineering}, pp. 293-–312.

\bibitem{Valero2009-2}
Valero, V., Cambronero, M.E., D\'{\i}az, G. and Maci\`a, H. (2009). A Petri net approach for the design and analysis ofWeb Services Choreographies, \emph{The Journal of Logic and Algebraic Programming} \textbf{78}, pp. 359-–380.

\bibitem{Verma1999}
Verma, D. (1999). Supporting Service Level Agreements on IP Networks, \emph{Macmillan Technical Publishing}.

\bibitem{W3C1999}
W3C (1999). XML Path Language (XPath) Version 1.0, \url{http://www.w3.org/TR/xpath/}.

\bibitem{W3C1999-2}
W3C (1999). XSL Transformations (XSLT) Version 1.0, \url{http://www.w3.org/TR/xslt}.

\bibitem{W3C2001}
W3C (2001). Web Service Definition Language (WSDL 1.1), \url{http://www.w3.org/TR/wsdl}.

\bibitem{W3C2002}
W3C (2002). Web Service Choreography Interface (WSCI) 1.0, \url{http://www.w3.org/TR/wsci/}.

\bibitem{W3C2004}
W3C (2004). Web Services Glossary, \url{http://www.w3.org/TR/ws-gloss/}.

\bibitem{W3C2004-2}
W3C (2004). OWL-S: Semantic Markup for Web Services, \url{http://www.w3.org/Submission/OWL-S/}.

\bibitem{W3C2005}
W3C (2005). Web Services Choreography Description Language Version 1.0, \url{http://www.w3.org/TR/ws-cdl-10/}.

\bibitem{W3C2006}
W3C (2006). Extensible Stylesheet Language (XSL) Version 1.1, \url{http://www.w3.org/TR/xsl/}.

\bibitem{W3C2007}
W3C (2007). SOAP Specifications, \url{http://www.w3.org/TR/soap/}.

\bibitem{Wing1990}
Wing, J.M. (1990). A Specifier's Introduction to Formal Methods, \emph{IEEE Computer}, \textbf{23}, pp. 8-–24.

\bibitem{Wright1999}
Wright, G.H.V. (1999). Deontic logic: A personal overview, \emph{Ratio Juris}, \textbf{12}, pp. 26-–38.

\bibitem{Yang2008}
Yang, H., Zhao, X., Cai, C. and Qiu, Z. (2008). Model-Checking of Web Services Choreography, \emph{Proceedings of Fourth IEEE International Symposium on Service-Oriented System Engineering}, pp. 79--84.

\bibitem{Zhao2003}
Zhao, J. and Rinard, M.C. (2003). Pipa: A behavioral interface specification language for AspectJ, \emph{Proceedings of Fundamental Approaches to Software Engineering} \textbf{2621}, pp. 150--165.

\end{thebibliography}