\section{Soundness for Continuous Time Workflow Nets} \label{sec:cont}

We have so far considered only discrete time workflow nets as this
is often sufficient for the practical applications. Nevertheless,
we shall now also discuss how the soundness
notion changes under the assumption of continuous time. 
For continuous time workflow nets,
we have marking as a function $M : P \longrightarrow \mathcal{B}(\rnul)$ 
where $\mathcal{B}(\rnul)$ is the set of all finite multisets over
nonnegative real numbers. Hence for every place $p \in P$ and 
every token $x \in M(p)$ we have $x \in \rnul$. We allow delay
transitions not only for integers but for arbitrary real numbers.
Otherwise the definition of continuous timed-arc workflow nets
is the same as in the case of the discrete semantics given earlier.

It is well-known that for closed 
timed automata (without any strict inequalities in guards), 
the location reachability problems for continuous and
discrete time delays coincide (see e.g.~\cite{AMP:CONCUR:98,danny_VLiet}).
We prove the same result also for extended timed-arc Petri
nets, following the idea from~\cite{danny_VLiet} where
the problem is reduced to an instance of 
linear programming, though with additional technical challenges 
for the Petri net case.
The result also implies that
%that the sets of untimed reachable markings
%in the continuous and discrete semantics coincide. Moreover, 
the discrete time semantics is sufficient for finding the minimum
and maximum execution times.

\begin{theorem}\label{thm:discvscont}
Let $N$ be an ETAPN and let $M_0$ be a marking on $N$ with all tokens
of integer ages. For any computation 
$M_0 \trans{d_0,t_0} M_1 \trans{d_1,t_1} M_2 \trans{d_2,t_2} 
\ldots \trans{d_{n-1},t_{n-1}} M_{n}$ in $N$ 
with $d_i \in \rnul$ for all $i$, $0 \leq i < n$, there is a computation 
$M_0 \trans{d^\prime_0,t_0} M^\prime_1 \trans{d^\prime_1,t_1} M^\prime_2 \trans{d^\prime_2,t_2} 
\ldots \trans{d^\prime_{n-1},t_{n-1}} M^\prime_{n}$ in $N$ such that
$d^\prime_i \in \nnul$ for all $i$, $0 \leq i < n$, and
$|M_n(p)|=|M^\prime_n(p)|$ for all $p \in P$.
Moreover if $N$ is a workflow net then there exists a computation with 
integer delays that achieves
the minimum, and if it exists also the maximum, execution time. 
\end{theorem}
\begin{proof}
Let $N$ be an extended timed-arc Petri net and let $M_0$ be its marking
with integer ages of tokens only.

Let $r$ be an execution 
\begin{equation}\label{eq:execution}
M_0 \trans{d_0,t_0} M_1 \trans{d_1,t_1} M_2 \trans{d_2,t_2} 
\ldots \trans{d_{n-1},t_{n-1}} M_{n}
\end{equation}
where $d_i \in \rnul$ for all $i$, $0 \leq i <n$.
Let $m=max\{|M_i| \text{ s.t. } 0 \leq i \leq n\}$ be the maximum number of
tokens in any of the intermediate markings.

In order to prove that the same sequence of transitions can be executed
also with integer time delays, we will first define a table
with $m$ rows and $n+1$ columns that represents how the tokens
in the net (rows) are moved around in the net by each transition firing
(column $i$ represents the marking $M_i$ that enables the firing of
transition $t_i$ and column $i+1$ the marking $M_{i+1}$ reached after its firing). 
The timing information is forgotten in the table, it merely represents
the consumption of tokens and whether they were produced by normal or
transport arcs. Based on such a table we will later define a difference
constraint system describing all possible delays that enable the firing
of the given transition sequence.

A table \itn{T} for the execution $r$ is a matrix with $m$ rows and
$n+1$ columns such that each element \tai \ of the table, where 
$1 \leq y \leq m$ and $0 \leq i \leq n$, contains either the value 
$\bot$ (unused token) or the pair $(p,f)$ where $p \in P$ represents the location of the token
and $f \in \nnul \cup \{\bullet\}$ is a flag signalling whether
the age of the given token was reset to some given value from $\nnul$
or whether it has not changed (the flag value $\bullet$).
We denote by \taip and \taif
the elements $p$ and $f$ of the pair of \tai, respectively.
If one element of the pair $(p,f)$ is not relevant for our considerations,
we simply write $(p,-)$ or $(-,f)$. We let $y$ to range over
the rows in the table (tokens) and $i$ over the columns in the table
(transition firing steps).

\begin{definition}[Valid table for run $r$] 
A table $T$ for the run $r$ is \emph{valid} if the following conditions
are met.
\begin{description}
\item[a)] Given the initial marking 
$M_0=\{(p_1,x_1),(p_2,x_2),\ldots,(p_k,x_k)\}$, 
the zero column of \itn{T} is defined as
%consists of \itn{T} includes a column \taj{y}{0} such that 
$\taj{y}{0}\eqdef(p_y,x_y)$ if $1\leq y \leq k$, and 
$\taj{y}{0}\eqdef\bot$ if $k <  y \leq m$.
\item[b)] For each column \itn{i} in the table:
\begin{itemize}
\item there is a set $\consumei\subseteq \jindex$ representing 
the $y$-indexes of the tokens  in column $i$
consumed by firing the transition $t_i$ such that for all 
$p \in {^\bullet}t_i$ we have 
$w(p,t_i)=|\{y \in \consumei \mid \taip=p\}|$,
and for all $p \in P \setminus \!^\bullet{t_i}$ we have 
$|\{y \in \consumei \mid \taip=p\}|=0$, 
\item there is a set $\producei\subseteq \jindex$ representing
the $y$-indexes of the tokens in column $i+1$ produced by firing 
the transition $t_i$ such that for all 
$p \in t_i^\bullet$ we have 
$w(t_i,p)=|\{y \in \producei \mid \tajp{y}{i+1}=p\}|$ 
and for all $p \in P\,\setminus\, t_i^\bullet$ we have 
$|\{y \in \producei \mid \tajp{y}{i+1}=p\}|=0$, and
\item there is a bijection 
$\itn{\mathscr{P}}:\jindex\rightarrow\jindex$ that relates the indexes 
of column $i$ with 
those presented in column $i+1$ such that
\begin{enumerate}[i)]
\item if $|\consumei| \leq |\producei|$ and $y \in \consumei$ then $\mathscr{P}(y) \in \producei$,
\item if $|\consumei| \geq |\producei|$ and $\mathscr{P}(y) \in \producei$ then $y \in \consumei$,
\item if $y \in \consumei$ and 
$\type((\taip,t_i))=\transporti= \type((t_i,p'))$ 
then $\mathscr{P}(y)=y$ and $\tajp{y}{i+1}=p'$,
\item if $y \in \jindex \setminus \consumei$ and $\tai\neq\bot$ then $\mathscr{P}(y)=y$ and $\tajp{y}{i+1}=\taip$,
\item if  $y \in \jindex\setminus \consumei$ and $\tai=\bot$ then either $\mathscr{P}(y) \in \producei$, or $\mathscr{P}(y)=y$ and $\taj{y}{i+1}=\bot$, and

%\item if $a \in \producei$ and $\type(t_i,\taip)=Transport_l$ and $\type(p^\prime,t_i)=Transport_l$ for some l then $\mathscr{P}(a)=a$,
\item if $T_{\mathscr{P}(y),i+1}=\bot$ then $y \in \consumei$ or $\tai=\bot$.
\end{enumerate}
\end{itemize}

\item[c)] For each column $i$ in the table:
\begin{itemize}
\item if $\type((p,t_i))=\inhib$ for some $p \in P$ 
then for all $y \in \jindex$, $\taip\neq p$,
\item for all $y \in \producei$, if
${\mathit \type((t_i,T^{place}_{y,i+1}))=\normal}$ 
then ${\mathit T^{flag}_{y,i+1}=0}$ else ${\mathit T^{flag}_{y,i+1}=\bullet}$,
and
%(${\mathit \type((t_i,T^{place}_{y,i+1}))=Transport_\ell}$ for some $\ell$), 
\item if $y \notin \producei$ and $\taj{y}{i+1} \neq \bot$ then ${\mathit T^{flag}_{y,i+1}=\bullet}$.
\end{itemize}
\end{description}
\end{definition}
Let us now assume a  computation 
$$M_0 \trans{0,start} M_1 \trans{3.5,book} M_2 \trans{4.6,pay}  
M_3 \trans{0,empty} M_4 \trans{0,empty}  M_5 \trans{0,empty} 
M_6 \trans{0,success}M_7$$ 
in our running example from Figure~\ref{fig:example}, where $M_0=\{(in,0)\}$.
One possible valid table $T$ for this computation is given
in Table~\ref{validtable}.

\begin{table}
\scriptsize
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
 \hline
 & $M_0$ & $M_1$ & $M_2$ & $M_3$ & $M_4$ & $M_5$ & $M_6$ & $M_7$\\
\hline
1 & (in,0) & (booking,0)   & (payment,$\bullet$) & (successful,0) & (successful,0) & (successful,0) & (successful,0) & (out,0)\\\hline
2 & $\bot$   & (attempts,0) & (attempts,$\bullet$) & (attempts,$\bullet$) & (attempts,$\bullet$) & (attempts,$\bullet$) & $\bot$ & $\bot$\\\hline
3 & $\bot$   & (attempts,0) & (attempts,$\bullet$) & (attempts,$\bullet$) & (attempts,$\bullet$) & $\bot$ & $\bot$ & $\bot$\\\hline
4 & $\bot$   & (attempts,0) & (attempts,$\bullet$) & (attempts,$\bullet$) & $\bot$ & $\bot$ & $\bot$ & $\bot$\\
\hline
\end{tabular}
\vspace{3mm}
\caption{A valid table for this run.}
\vspace{-3mm}
\label{validtable}
\end{table}

One can verify that there is a valid table for any computation of the
given net $N$. On the other hand, any valid table defines a legal computation 
on untimed markings represented by each column.

\begin{definition}[Untimed marking given by column $i$] \label{def:untmark}
Let $T$ be a valid table and let $0 \leq i < n$.
We define the untimed marking $M_i^u \eqdef \{ \tajp{y}{i} \in P \mid
1 \leq y \leq m  \}$  as a multiset of all places where a token is present
in the column $i$ of the table $T$.
\end{definition}

By the way a valid table is constructed, the reader can easily verify
the validity of the following lemma.
\begin{lemma}[Untimed consistency of a valid table $T$]\label{lemma:untmarking}
Let \itn{T} be a valid table.
Then $M^u_i \trans{t_i} M^{u}_{i+1}$ for all $i$, $0 \leq i < n$,
in the classical (untimed) Petri net semantics.
\end{lemma}

We shall now proceed with defining a set of difference constraint 
inequalities encoding in a sound and complete way 
the timings aspects of the computation given in Equation~(\ref{eq:execution}).

Let the \emph{execution time} of a transition $t_i$
in Equation~(\ref{eq:execution}) be denoted by the variable $e_i$
representing the total time elapsed from the 
initialization until the transition $t_{i}$ is fired. 
In order to construct the system of inequalities over the variables
$e_0, \ldots, e_{n-1}$, we need to define an expression describing
the age of a token $y$ just at the moment the transition $t_i$ is fired.

\begin{definition}[Token-age expression]\label{def:age}
Let $T$ be a valid table.
We define \agexi{y}{i} where $1 \leq y \leq m$ and $0 \leq i < n$
as the expression 
\begin{center}
``$e_i -e_{j-1} +d$'' 
\end{center}
where $j$, $j\leq i$,
is the largest number not greater than $i$
% entry time of the marking that contains the first element of T 
such that $\tajf{y}{j}=d\in \nnul$. 
%By definition, $e_{-1}=\agexi{y}{-1}=0$ for all $y \in \jindex$.
\end{definition}

The intuition is that $\agexi{y}{i}$ expresses, in terms of
the execution time variables, the current age
of token $y$ after the time delay $d_i$ and exactly when 
firing the transition $t_i$.
The correctness of the definition follows from the
requirements on a valid table and the fact that the first column (i=0) of 
any valid table contains only pairs $(p,f)$ where $f \not= \bullet$.

For example, in our running example in Table~\ref{validtable},
the age of token $1$ at the moment the transition \itn{pay} is fired
can be expressed as $e_2 - e_0 + 0$.


%{\bf Remark.} It is important to note that $\agexi{y}{i}$ always exists for all $y \in \jindex$ 
%and $i \in \{-1,0,\ldots,n\}$ since if $\type((t_{i-1}\taip))=\normal$  then
%$\taj{y}{i}=(-,0)$ and, therefore, $j=i$, whereas if $\type((t_{i-1}\taip))=Transport_\ell$ for some $\ell$,
%we will always find such a $j$ since, according to the definition of the bijective function $\mathscr{P}$ (condition iii)) and
%the fact that the multiplicity of the output arc of this transport arc is by definition 1, one can always search backward in the same
%row and find, at least, one element such that $\taif\neq \bullet$.   

We are now ready to define, for a given valid table, a system of inequalities
over the variables $e_i$ that expresses all the timing constraints on the 
firing of transitions, age invariants and urgency.
 
\begin{definition}[Constraint system]\label{def:consys}
Let \itn{T} be a valid table for a run $r$ from
Equation~(\ref{eq:execution}). The constraint system $\mathscr{C}$ for $T$ 
is the set of inequations over the variables
$e_0, e_1, \ldots, e_{n-1}$, containing the constraints 
$\{ e_0 \leq e_1,\ e_1 \leq e_2, \ldots, e_{n-2} \leq e_{n-1}\}$, 
and constructed 
such that for all $p \in P$, $y \in \jindex$ and $i \in \iindex$:
\begin{description}
\item[a)] if $\taip=p$ and $\inv(p)=[0,u]$ where $u \in \nnul$, 
we add the inequality $\itn{age(y,i)\leq u}$ to $\mathscr{C}$,
\item[b)] if $\taip=p$ and $y \in \consumei$ and $(p,t_i) \in \ia$ and $\cfunction((p,t_i))=[a,b]$, we add $a \leq \agexi{y}{i}$ and if $b \not=\infty$ also 
$\agexi{y}{i}\leq b$ to $\mathscr{C}$, and
\item[c)] if $M^u_i$ enables some $t \in \Turg$
then we add $e_i - e_{i-1} = 0$ to $\mathscr{C}$
where by definition $e_{-1}$ is replaced with $0$.
\end{description}
\end{definition}

In our running example (Figure \ref{fig:example}), the constraint system for the valid table
depicted in Table~\ref{validtable} is the following.
\begin{itemize}
\item We add the inequalities
$$e_0 \leq e_1,\ \ \  e_1 \leq e_2,\ \ \ \ldots,\ \ \ e_5 \leq e_6\ .$$
\item For the two nontrivial age invariants, we add
$\agexi{1}{1} \leq 5$ and $\agexi{1}{2} \leq 10$, in other words
we add the constraints
$$e_1-e_0 \leq 5,\ \ \  e_2-e_0 \leq 10 \ . $$
\item Regarding the guards on input arcs, we add for column $0$
the constraint $0 \leq\agexi{1}{0}$, for column $1$ the constraints
$2 \leq \agexi{1}{1}$ and
$\agexi{1}{1} \leq 5$, for column $2$ the constraints
$0 \leq \agexi{1}{2}$ and
$\agexi{1}{2} \leq 10$ and so on for the remaining columns (producing
only trivial constrains that are always satisfied). Hence the following
constraints (listing only the nontrivial ones) are added to the systems:
$$2 \leq e_1 - e_0,\ \ \ e_1 - e_0 \leq 5,\ \ \ e_2-e_0 \leq 10\ .$$
\item Finally, for each column that enables some urgent transition 
(columns $0$, $3$, $4$, $5$ and $6$), we add (recall that $e_{-1}=0$):
$$e_0=0, \ \ \ e_3=e_2, \ \ \ e_4=e_3, \ \ \ e_5=e_4, \ \ \ e_6=e_5\ .$$
\end{itemize}

Observe that the original delays in the trace from Equation~(\ref{eq:execution})
form a solution of the constructed constraint system: 
$e_0=0$,
$e_1=3.5$,
$e_2=8.1$,
$e_3=8.1$,
$e_4=8.1$,
$e_5=8.1$,
$e_6=8.1$.
In fact, there is also an integer solution to the constraint system
(this is not only a coincidence),
e.g.
$e_0=0$,
$e_1=2$,
$e_2=3$,
$e_3=3$,
$e_4=3$,
$e_5=3$,
$e_6=3$,
and such a sequence is executable in our running workflow example.

\begin{lemma} \label{lem:correct}
Let $r$ be a run $M_0 \trans{d_0,t_0} M_1 \trans{d_1,t_1} M_2 \trans{d_2,t_2} 
\ldots \trans{d_{n-1},t_{n-1}} M_{n}$ with $d_i \in \rnul$ in a workflow
net $N$. Then there is a valid table $T$ for $r$
and the corresponding constraint system $\mathscr{C}$ such that
$e_i= \sum_{j=0 \ldots i} d_j$ is a solution of $\mathscr{C}$. 
Moreover, $e_0, e_1, \ldots, e_{n-1}$ is a (real) solution of 
$\mathscr{C}$ if and only if
$M_0 \trans{e_0,t_0} M_1 \trans{e_1-e_0,t_1} M_2 \trans{e_2-e_1,t_2} 
\ldots \trans{e_{n-1}-e_{n-2},t_{n-1}} M_{n}$.
\end{lemma}
\begin{proof}[sketch]
By analysing the requirements for a valid table, we can see that if
a run $r$ can be performed in the net $N$ then we are able to 
design a table that satisfies all the requirements for the untimed
part of the run execution and uses the right tokens in the pairing
bijection such that the corresponding constraint system $\mathscr{C}$
gives the sufficient and necessary
conditions for the execution time variables $e_i$ to produce a valid
computation of the net $N$ with the given sequence of transitions firing. 
Hence the original execution times in the given run are one possible solution of 
the system but any such a solution actually provides a possible
timed execution of the run (note that if a transition $t_i$ is performed 
at time $e_i$ and transition $t_{i+1}$ is performed at time $e_{i+1}$ then
the delay between the execution of these two transitions is $e_{i+1}-e_{i}$).
\qed
\end{proof}

We can now summarise and conclude the proof of Theorem~\ref{thm:discvscont}. 
We assumed a run of the net with real delays as in Equation~(\ref{eq:execution}).
Based on this we know that there exists a valid table $T$ for such a run such that 
the constraint system $\mathscr{C}$ for the run, representing all possible 
delays that can execute the transition sequence in~(\ref{eq:execution}), has
a solution corresponding to the delays in (\ref{eq:execution}). This is due
to Lemma~\ref{lem:correct}. 

The constraint system $\mathscr{C}$ is an instance of linear programming 
problem where we used difference constraints only and we are therefore 
guaranteed that the matrix in the linear programming problem is totally
unimodular~\cite{umodmatrix1}. As the system has a solution, it also
has an optimal integral solutions---a general result guaranteed
for any linear programming problem with totally unimodular 
matrix~\cite{umodmatrix2}.
Hence using Lemma~\ref{lem:correct} we know that there is also an execution
of $N$ following the same transitions as in Equation~(\ref{eq:execution})
but with integral delays only. Moreover, the minimum and maximum
execution times can be also achieved by integral delays only. Hence
the proof of Theorem~\ref{thm:discvscont} is completed.

\end{proof}

As expected, continuous soundness now implies soundness in the discrete case.
Moreover, the discrete and continuous soundness coincide for a subclass
of workflow nets that do not enforce any urgent behaviour.

\begin{theorem} \label{thm:implication}
Let $N$ be an ETAWFN. If $N$ is sound in the continuous semantics
then it is sound in the discrete semantics.
\end{theorem}
\begin{proof}
Let $N$ be sound in the continuous semantics. Let $M$ be a marking
reachable from the initial marking $M_\mathit{in}$ in the discrete semantics.
As $M$ is clearly reachable also in the continuous semantics,
condition $b)$ of Definition~\ref{def:soundness} is clearly satisfied.
Regarding condition $a)$ of the definition of soundness, we know
that some final marking $M_\mathit{out}$ is reachable from $M$ in the continuous
semantics. However, using Theorem~\ref{thm:discvscont} we can conclude
that a marking $M'_\mathit{out}$ that has the same distribution of tokens
as $M_\mathit{out}$ is reachable from $M$ also in the discrete semantics,
and hence $N$ is sound w.r.t. the discrete semantics.
\end{proof}

If we moreover consider only workflow nets where time delays are not restricted
by neither age invariants nor urgency, then both the continuous and discrete
semantics coincide with respect to soundness. 

\begin{theorem} \label{thm:contdiscok}
Let $N$ be an ETAWFN with no age invariants and no urgent transitions
(inhibitor arcs are allowed). Then $N$ is sound in the
continuous semantics if and only if $N$ is sound in the discrete semantics.
\end{theorem}
\begin{proof}
``$\Rightarrow$'': Follows from Theorem~\ref{thm:implication}.\\
\indent ``$\Leftarrow$'': Let $N$ be sound in the discrete semantics.
Let $M$ be a marking
reachable from the initial marking $M_\mathit{in}$ in the continuous
semantics.
%We can w.l.o.g. assume that $M_\mathit{in}$ is a marking
%where the token in the place \itn{in} has an integer age (if not then
%we start with the integer part of the token age and as the first
%transition delay until the fractional part is achieved; this is always
%possible as the net does not contain any age invariants or urgent transitions).
Now we can use Theorem~\ref{thm:discvscont} to argue that a marking $M'$
is reachable from $M_\mathit{in}$ in the discrete semantics (with integer
delays only) such that $|M(p)|=|M'(p)|$ for all $p \in P$.
As $M$ and $M'$ have the same number of tokens in all the places
and $N$ is sound in the discrete semantics, it follows that condition $b)$
in Definition~\ref{def:soundness} holds for $M'$ and hence also for $M$.
Let us now argue for condition $a)$.
Let $M \trans{d} M_1$ where $d$ is an integer
greater than any constant used in any interval on any input arc.
Such a delay is possible as the net does not contain
any age invariants or urgent transitions. Clearly, $M' \trans{d} M'_1$
is possible also in the discrete semantics. However, now
all tokens in $M_1$ and $M'_1$ are greater than any constant appearing
in the net and hence these two markings are timed bisimilar.
Because $N$ is sound in the discrete semantics, we know that
there is some final marking $M_\mathit{out}$ such that
$M_\mathit{out} \in [M'_1\rangle$. As $M_1$ is timed bisimilar with
$M'_1$, it can also reach a final marking (bisimilar to $M_\mathit{out}$)
and so does $M$. Hence condition $a)$ of Definition~\ref{def:soundness}
is established and we can conclude that
$N$ is sound also in the continuous semantics.
\end{proof}


\begin{figure}[!h]
%\subfloat[][Using invariants]{
\begin{minipage}[c]{0.5\textwidth}
\begin{tikzpicture}[font=\scriptsize,xscale=1.9,yscale=1.6]
  \tikzstyle{arc}=[->,>=stealth,thick]
	\tikzstyle{transportArc}=[->,>=diamond,thick]
  \tikzstyle{every place}=[minimum size=6mm,thick]
  \tikzstyle{every transition}=[fill=black,minimum width=2mm,minimum height=5mm]
  \tikzstyle{every token}=[fill=white,text=black]
  \tikzstyle{urgency}=[place,fill=white,minimum size=2.0mm,thin]
%        \node[] at (0,-0.8) (space) { };
	\tikzstyle{inhibArc}=[->,>=o,thick]
\node[place,label=above:$\itn{in}$,structured tokens={0},] at (0.6,0) (in) {};
\node[place,label=above:$\itn{out}$,] at (3.25,0) (out) {};
\node[place,label=above:$\itn{waiting}$,] at (1.2,1) (waiting) {};
\node[place,label=left:$\itn{deadline}$,label=below:inv: $\leq  1$,] at (1.2,-0.9) (deadline) {};
\node[place,label=above:$\itn{finished}$,] at (2.25,1) (finished) {};
\node[transition,label=right:$\itn{init}$] at (1.2,0) (init) {};
\node[transition,label=below:$\itn{service}$] at (1.75,1) (service) {};
\node[transition,label=below:$\itn{late}$] at (2.25,0) (late) {};
\node[transition,label=below:$\itn{early}$] at (2.75,-0.9) (early) {};
\draw[arc] (in) -- (init) node[midway,auto] {$$} {};
\draw[arc] (init) -- (waiting)  {};
\draw[arc] (init) -- (deadline)  {};
\draw[arc] (waiting) -- (service) node[midway,auto] {$$} {};
\draw[arc] (service) -- (finished)  {};
\draw[arc] (finished) -- (late) node[midway,auto] {$[0,0]$} {};
\draw[arc] (deadline) -- (late) node[midway,auto] {$$} {};
\draw[arc] (late) -- (out)  {};
\draw[arc] (deadline) -- (early) node[midway,auto] {$$} {};
\draw[arc] (finished) -- (2.75,1) -- (early) node[very near start,auto] {$[1,1]$} {};
\draw[arc] (early) -- (3.25,-0.9) -- (out)  {};
\end{tikzpicture}
\captionsetup{type=subfigure}
\caption{Using invariants.}
\label{fig:count1}
\end{minipage}
\hspace{0mm}
%\subfloat[][Using urgency]{
\begin{minipage}[c]{0.5\textwidth}
\begin{tikzpicture}[font=\scriptsize,xscale=1.9,yscale=1.6]
  \tikzstyle{arc}=[->,>=stealth,thick]
	\tikzstyle{transportArc}=[->,>=diamond,thick]
  \tikzstyle{every place}=[minimum size=6mm,thick]
  \tikzstyle{every transition}=[fill=black,minimum width=2mm,minimum height=5mm]
  \tikzstyle{every token}=[fill=white,text=black]
  \tikzstyle{urgency}=[place,fill=white,minimum size=2.0mm,thin]
	\tikzstyle{inhibArc}=[->,>=o,thick]
	\tikzstyle{urgency}=[place,fill=white,minimum size=2.0mm,thin]
 \node at (0,0) {};
\node[place,label=above:$\itn{in}$,structured tokens={0},] at (0,1) (in) {};
\node[place,label=below:$\itn{idle}$,] at (1.25,1) (P1) {};
\node[place,label=below:$\itn{out}$,] at (2.4,1) (out) {};
\node[transition,label=above:$\itn{start}$] at (0.65,1) (start) {};
\node[transition,label=above:$\itn{loop}$] at (1.25,2.55) (loop) {};
\node[urgency] at (1.25,2.55) {};
\node[transition,label=above:$\itn{end1}$] at (1.9,1) (end1) {};
\node[transition,label=above:$\itn{end2}$] at (2.4,2.55) (end2) {};
\draw[arc] (P1) -- (end1) node[midway,below,xshift=-1] {$[0,0]$} {};
\draw[arc] (P1) -- (end2) node[midway,above] {} {};
\draw[arc] (end1) -- (out)  {};
\draw[arc] (end2) -- (out)  {};
\draw[transportArc] (in) -- (start) node[midway,auto] {:1} {};
\draw[transportArc] (start) -- (P1) node[midway,auto] {:1} {};
\draw[transportArc] (P1) to [bend right=25] (loop) node[near start,auto] {};
\draw[transportArc] (loop) to [bend right=25](P1) node[near end,auto] {};
\node [] at (0.9,1.85) {:1};
\node [] at (1.6,2) {:1};
\node [] at (1.94,2.29) {$[1,\infty)$};
\end{tikzpicture}
\vspace{-0.4cm}
\captionsetup{type=subfigure}
\caption{Using urgency.}\label{fig:count2}
\end{minipage}
\caption{Sound nets in the discrete semantics and unsound in the continuous one}
\label{fig:counts}
\end{figure}

%We know by Theorem~\ref{thm:discvscont} that the
%same untimed markings are reachable both in the discrete and continuous
%semantics, however, 
For extended timed-arc 
workflow nets the notion of soundness for the discrete and continuous 
semantics are, perhaps surprisingly, different.
Consider Figure~\ref{fig:counts} where both given workflow nets are sound
in the discrete semantics but not in the continuous one.

%. However, in the continuous semantics we can execute in the net
%from Figure~\ref{fig:count1} the sequence
%``\itn{init}, delay $0.5$, \itn{service}, delay $0.5$'', bringing us
%into a deadlock situation. In the net from Figure~\ref{fig:count2},
%we can perform the sequence ``delay $0.5$, \itn{start}'', ending up
%in a situation where only the urgent transition
%\itn{loop} is enabled. Hence the nets are not sound w.r.t. the continuous semantics.

\begin{theorem} \label{thm:counterex}
There is an ETAWFN (with either age invariants
or urgent transitions) sound in the
discrete semantics but unsound in the continuous one.
\end{theorem}
\begin{proof}
Consider the nets in Figure~\ref{fig:counts} (as before we do not
draw the $[0,\infty]$ intervals). It is easy to verify that
both of them are sound w.r.t. the discrete semantics. Indeed,
in Figure~\ref{fig:count1} the age of the token in the place
\itn{finished} can be either $1$ or $0$, depending on whether the
service was executed early or late, and then either the transition
\itn{early} or \itn{late} will be enabled and allow us to reach a
final marking. Similarly in Figure~\ref{fig:count2} the age of the token
in the place \itn{idle} will be of integer age so even though the transition
\itn{loop} disables any time delay, we can still terminate the workflow
by firing either the transition \itn{end1} or \itn{end2}.

However, in the continuous semantics we can execute in the net
from Figure~\ref{fig:count1} the sequence
``\itn{init}, delay $0.5$, \itn{service}, delay $0.5$'', bringing us
into a deadlock situation. In the net from Figure~\ref{fig:count2},
we can perform the sequence ``delay $0.5$, \itn{start}'', ending up
in a situation where only the urgent transition
\itn{loop} is enabled but at the same time disallows for any time delay.
In both cases the nets are not sound w.r.t. the continuous semantics.
\end{proof}
